---
title: "566 Project Data Exploration"
author: "North Carpenter"
date: "2026-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(sf)
library(ggplot2)
```

# Importing Data

```{r}
# This block will import key data. It is imparative that you structure your
# files in the same way or this will not work. Set up this file in the 
# Y-crash-project folder somewhere on your computer and at the same level of the 
# project folder put the 566 project data folder.
#
# You should be able to import the files in 566 in this chunk only without issue.

# Importing Data

# Import any csv's here
accident2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/accident.csv")
person2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/PERSON.csv")
vehicle2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/VEHICLE.csv")
drdistract2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/distract.csv")
factor2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/factor.csv")
nmCrash2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/nmcrash.csv")
cEvent2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/cevent.csv")
vission2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/vision.csv")

# Just Curios
drugs2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/drugs.csv")
```

# Data Overiview

```{r}
looknon <- function(df)
{
  # Take a DF and returns a new df that counts na/"" in one column and 
  # "None"/"none" in another column. Also returns the number of unique values 
  # per column.
  
  df <- df |>
    mutate(across(everything(), as.character)) %>% #make everything char for ==
    pivot_longer(everything(), names_to = "column", values_to = "value") %>%
    group_by(column) %>% # set up groups
    summarise(
      num_missing = sum(is.na(value) | value == "", na.rm = TRUE),
      num_None = sum(value == "None" | value == "none", na.rm = TRUE),
      num_unique  = n_distinct(value)
    ) #summarize to these col
  return(df)
}
```

## Vission
```{r}
countVission <- vission2023 |>
  count(VISION)
countVission
sum(countVission$n) - 54793

3751 / sum(countVission$n)
```


## Vehicle Stats
```{r}
vehicle2023 |>
  count(M_HARM)

vehicle2023 |>
  count(VPAVETYPNAME)

vehicle2023 |>
  count(VSURCONDNAME)
```


## Acidedent Summary
```{r}
looknon(accident2023)

```

## veVent File
```{r}
head(veVent2023)

veVent2023 |>
  group_by(ST_CASE)|>
  summarise(n = n(), .groups = "drop")

veVent2023 |>
  group_by(SOE)|>
  summarise(n = n(), .groups = "drop")
```


## Pedestrians and Injuries
```{r}
person2023 |>
  group_by(PER_TYPNAME, INJ_SEVNAME) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(
    names_from  = INJ_SEVNAME,
    values_from = n,
    values_fill = 0
  )
```

## Person Type
```{r}
person2023 |>
  count(PER_TYPNAME)

person2023 |>
  group_by(PER_TYPNAME, DRINKINGNAME) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(
    names_from  = DRINKINGNAME,
    values_from = n,
    values_fill = 0
  )
```


## Alcohol
```{r}
person2023 |>
  count(ALC_RESNAME)

person2023 |>
  count(DRINKINGNAME)

person2023 |>
  group_by(PER_TYPNAME, DRINKINGNAME) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(
    names_from  = DRINKINGNAME,
    values_from = n,
    values_fill = 0
  )
```

## Drugs
```{r}
drugtype <- drugs2023 |> 
  group_by(DRUGRESNAME) |> 
  summarise(n = n()) |> 
  mutate(percent = n / sum(n))
drugtype

drugs2023 |>
  count(DRUGRESNAME)
```

# Pretty Pictures
```{r}
accident2023noLat <- accident2023 |>
  filter(STATENAME != "Alaska",
         LATITUDE > 24, LATITUDE < 72,
         LONGITUD > -180, LONGITUD < -65)


my_sf <- st_as_sf(accident2023noLat, coords = c('LONGITUD', 'LATITUDE'))

my_sf <- st_set_crs(my_sf, 4326)

#Plot it:

ggplot(my_sf) + 
  geom_sf(aes(color = STATENAME), size = 0.01) +
  theme(legend.position = "none")

```

```{r}
accident2023seattle <- accident2023 |>
  filter(STATENAME == "Connecticut",
         LATITUDE > 24, LATITUDE < 72,
         LONGITUD > -180, LONGITUD < -65)

axSeattleMap <- st_as_sf(accident2023seattle, coords = c('LONGITUD', 'LATITUDE'))

axSeattleMap <- st_set_crs(axSeattleMap, 4326)

#Plot it:

ggplot(axSeattleMap) + 
  geom_sf(aes(color = FATALS), size = 1)

```

# Cleaning
## Setting Up Variables to Keep
```{r}
merging_Vars <- c("ST_CASE", "VEH_NO", "PER_NO")
accident_keep <- c("VE_FORMS", "MONTH", "DAY", "YEAR", "HOUR", "MINUTE", "DAY_WEEK", "ROUTE", "RUR_URB", "RELJCT1", "LGT_COND", "WEATHER", "ARR_HOUR", "ARR_MIN")
vehicle_keep <- c("TRAV_SP", "DR_DRINK", "PREV_ACC", "VSPD_LIM", "VPAVETYP", "VSURCOND", "P_CRASH1", "M_HARM")
person_keep <- c("AGE", "SEX", "PER_TYP", "INJ_SEV", "DRUGS", "DEATH_MO", "DEATH_DA", "DEATH_YR", "DEATH_TM", "DEATH_HR", "DEATH_MN", "LAG_HRS", "LAG_MINS") #Record Sex of Driver Only, fix type, fix type, complex death after indicator
vision_keep <- c("VISION")
fact_keep <- c("VEHICLECC")
dist_keep <- c("DRDISTRACT")
NMCC_keep <- c("NMCC")
```

## Triming Files
```{r}
data_list <- list(accident2023, vehicle2023, person2023, factor2023, drdistract2023, nmCrash2023, vission2023)
var_list <- list(accident_keep, vehicle_keep, person_keep, fact_keep, dist_keep, NMCC_keep, vision_keep)

data_list <- Map(function(df, vars){
  df %>% select(any_of(c(merging_Vars, vars)))
}, data_list, var_list)
```

## Check Worked
```{r}
lapply(data_list, head)
```

## Joining Non-Vehicle Files
```{r}
crashes2023 <- left_join(data_list[[3]], data_list[[1]], by = join_by(ST_CASE))

# in 2023 2 pedestrians were never killed in the same accident
crashes2023 |>
  group_by(ST_CASE, PER_TYP, INJ_SEV) |>
  summarise(n = n(), .groups = "drop") |>
  filter(PER_TYP == 5, INJ_SEV == 4) |>
  arrange(desc(n))

vehicle2023 |> 
  count(ST_CASE) |> 
  arrange(desc(n))

crashes2023 <- left_join(crashes2023, data_list[[6]], by = join_by(ST_CASE, PER_NO, VEH_NO))
```

## Dealing with Vehicles

The person level data is associate with the vehicle the person was in at the time of the accident. For pedestrians that means they are always in their own vehicle with a bunch of n/as listed. We need each pedestrian file incident to be associated with the vehicle that killed/injured them. In doing so we need to list vehicles for other accidents or have a bunch of perfectly correlated variables and loose useful information for persons in other accidents. For accidents with pedestrians I will join the vehicle with "most harmful event" listed as "pedestrian" for all others I will like the first vehicle to act in the sequence of events listed in veVent.

Our hierarchy is simply ordered 

```{r}
# Get vehicle number for first vehicle to act in the event order
acLVLcEvent <- pivot_wider(cEvent2023, names_from = EVENTNUM, values_from = VNUMBER1)
acLVLcEvent <- acLVLcEvent %>% filter(!is.na(`1`))
acLVLcEvent <- acLVLcEvent |> rename('vEvent1' = '1')
acLVLcEvent <- acLVLcEvent |> subset(select = c('ST_CASE', 'vEvent1'))
head(acLVLcEvent)

# Dropping all rows not the initiating vehicle
crashInishator <- left_join(data_list[[2]], acLVLcEvent, by = 'ST_CASE')
crashInishator <- crashInishator %>% filter(vEvent1 == VEH_NO)

# Join other vehicle factors to sparce vehicle matrix
crashInishator %>%
  

```

