---
title: "566 Project Data Exploration"
author: "North Carpenter"
date: "2026-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(sf)
library(ggplot2)
```

# Importing Data

```{r}
# This block will import key data. It is imparative that you structure your
# files in the same way or this will not work. Set up this file in the 
# Y-crash-project folder somewhere on your computer and at the same level of the 
# project folder put the 566 project data folder.
#
# You should be able to import the files in 566 in this chunk only without issue.

# Importing Data

# Import any csv's here
accident2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/accident.csv")
person2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/PERSON.csv")
vehicle2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/VEHICLE.csv")
drdistract2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/distract.csv")
factor2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/factor.csv")
nmCrash2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/nmcrash.csv")
cEvent2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/cevent.csv")
vission2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/vision.csv")

# Just Curios
drugs2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/drugs.csv")
```

# Data Overiview

```{r}
looknon <- function(df)
{
  # Take a DF and returns a new df that counts na/"" in one column and 
  # "None"/"none" in another column. Also returns the number of unique values 
  # per column.
  
  df <- df |>
    mutate(across(everything(), as.character)) %>% #make everything char for ==
    pivot_longer(everything(), names_to = "column", values_to = "value") %>%
    group_by(column) %>% # set up groups
    summarise(
      num_missing = sum(is.na(value) | value == "", na.rm = TRUE),
      num_None = sum(value == "None" | value == "none", na.rm = TRUE),
      num_unique  = n_distinct(value)
    ) #summarize to these col
  return(df)
}
```

## Vission
```{r}
countVission <- vission2023 |>
  count(VISION)
countVission
sum(countVission$n) - 54793

3751 / sum(countVission$n)
```


## Vehicle Stats
```{r}
vehicle2023 |>
  count(M_HARM)

vehicle2023 |>
  count(VPAVETYPNAME)

vehicle2023 |>
  count(VSURCONDNAME)
```


## Acidedent Summary
```{r}
looknon(accident2023)
```

## veVent File
```{r}
head(cEvent2023)

cEvent2023 |>
  group_by(ST_CASE)|>
  summarise(n = n(), .groups = "drop")

cEvent2023 |>
  group_by(SOE)|>
  summarise(n = n(), .groups = "drop")
```


## Pedestrians and Injuries
```{r}
person2023 |>
  group_by(PER_TYPNAME, INJ_SEVNAME) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(
    names_from  = INJ_SEVNAME,
    values_from = n,
    values_fill = 0
  )
```

## Person Type
```{r}
person2023 |>
  count(PER_TYPNAME)

person2023 |>
  group_by(PER_TYPNAME, DRINKINGNAME) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(
    names_from  = DRINKINGNAME,
    values_from = n,
    values_fill = 0
  )
```


## Alcohol
```{r}
person2023 |>
  count(ALC_RESNAME)

person2023 |>
  count(DRINKINGNAME)

person2023 |>
  group_by(PER_TYPNAME, DRINKINGNAME) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(
    names_from  = DRINKINGNAME,
    values_from = n,
    values_fill = 0
  )
```

## Drugs
```{r}
drugtype <- drugs2023 |> 
  group_by(DRUGRESNAME) |> 
  summarise(n = n()) |> 
  mutate(percent = n / sum(n))
drugtype

drugs2023 |>
  count(DRUGRESNAME)
```

# Pretty Pictures
```{r}
accident2023noLat <- accident2023 |>
  filter(STATENAME != "Alaska",
         LATITUDE > 24, LATITUDE < 72,
         LONGITUD > -180, LONGITUD < -65)


my_sf <- st_as_sf(accident2023noLat, coords = c('LONGITUD', 'LATITUDE'))

my_sf <- st_set_crs(my_sf, 4326)

#Plot it:

ggplot(my_sf) + 
  geom_sf(aes(color = STATENAME), size = 0.01) +
  theme(legend.position = "none")

```

```{r}
accident2023seattle <- accident2023 |>
  filter(STATENAME == "Connecticut",
         LATITUDE > 24, LATITUDE < 72,
         LONGITUD > -180, LONGITUD < -65)

axSeattleMap <- st_as_sf(accident2023seattle, coords = c('LONGITUD', 'LATITUDE'))

axSeattleMap <- st_set_crs(axSeattleMap, 4326)

#Plot it:

ggplot(axSeattleMap) + 
  geom_sf(aes(color = FATALS), size = 1)

```

# Cleaning
## Setting Up Variables to Keep
```{r}
merging_Vars <- c("ST_CASE", "VEH_NO", "PER_NO")
accident_keep <- c("VE_FORMS", "MONTH", "DAY", "YEAR", "HOUR", "MINUTE", "DAY_WEEK", "ROUTE", "RUR_URB", "RELJCT1", "LGT_COND", "WEATHER", "ARR_HOUR", "ARR_MIN")
vehicle_keep <- c("TRAV_SP", "DR_DRINK", "PREV_ACC", "VSPD_LIM", "VPAVETYP", "VSURCOND", "P_CRASH1", "M_HARM")
person_keep <- c("AGE", "SEX", "PER_TYP", "INJ_SEV", "DRUGS", "DEATH_MO", "DEATH_DA", "DEATH_YR", "DEATH_TM", "DEATH_HR", "DEATH_MN", "LAG_HRS", "LAG_MINS") #Record Sex of Driver Only, fix type, fix type, complex death after indicator
vision_keep <- c("VISION")
fact_keep <- c("VEHICLECC")
dist_keep <- c("DRDISTRACT")
NMCC_keep <- c("NMCC")
```

## Triming Files
```{r}
data_list <- list(accident2023, vehicle2023, person2023, factor2023, drdistract2023, nmCrash2023, vission2023)
var_list <- list(accident_keep, vehicle_keep, person_keep, fact_keep, dist_keep, NMCC_keep, vision_keep)

data_list <- Map(function(df, vars){
  df %>% select(any_of(c(merging_Vars, vars)))
}, data_list, var_list)
```

## Check Worked
```{r}
lapply(data_list, head)
```

## Creating Variables For Joins
Dealing with necessary variable manipulation prior to joining. Some variables have multiple entries for the unit they represent. Here I investigate that and transform the variables so the sets can be merged. See "Created Variable Dictionary" Google doc for a condensed description of what is done below.
### NMCrash
```{r}
data_list[[6]] %>% # There are multiple entries per combo of case, person, vehicle
  group_by(ST_CASE, PER_NO, VEH_NO) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n)) %>%
  count(n)

data_list[[6]] %>% # base frequency in unpivoted dataset
  count(NMCC)

wideNM <- data_list[[6]] %>% # Pivoting data set to have unique columns and correct number of rows.
  mutate(NMCC = if_else(NMCC == 0, 101, NMCC)) %>%
  pivot_wider(
    id_cols = c(ST_CASE, PER_NO, VEH_NO),
    names_from = NMCC,
    values_from = NMCC,
    values_fill = 0,
    values_fn = max
  )

uniteCount <- wideNM %>% # creating a data set of unique combos
  mutate(flag = if_else(
    rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(2, 3, 11, 12))) > 0, 1, 0
  )) %>%
  unite("combo", -c(ST_CASE, PER_NO, VEH_NO, flag), sep = "_") %>%
  count(combo, flag, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100,
         cum_pct = cumsum(pct))

uniteCount %>%
  filter(flag == 1) %>%
  summarise(total_pct = sum(pct))

# DEFFINITIONS OF NEW VARIABLES
wideNM <- wideNM %>%
  mutate(
    NMCCSimple = case_when(
      rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(101))) > 0 ~ 0,
      rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(98, 99))) > 0 ~ 5,
      rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(2))) > 0 ~ 1,
      rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(19))) > 0 ~ 3,
      rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(2, 3, 4, 11, 12))) > 0 ~ 3,
      TRUE ~ 4
    )
  )

# Drop Unused columns
wideNM <- wideNM %>%
  select(c(ST_CASE, PER_NO, VEH_NO, NMCCSimple))
```

### Initiating Vehicle Identification
Making a list of the vehicle that initiate each accident.
```{r}
# Get vehicle number for first vehicle to act in the event order
acLVLcEvent <- pivot_wider(cEvent2023, names_from = EVENTNUM, values_from = VNUMBER1) %>%
  filter(!is.na(`1`)) %>%
  rename('vEvent1' = '1') %>%
  subset(select = c('ST_CASE', 'vEvent1'))
head(acLVLcEvent)

# Dropping all rows not the initiating vehicle
crashInishator <- left_join(data_list[[2]], acLVLcEvent, by = 'ST_CASE') %>%
   filter(vEvent1 == VEH_NO) %>% # Keep only first vehicles
  select(-vEvent1) # Drop indicator now equal to VEH_NO
head(crashInishator)

nrow(crashInishator)
nrow(accident2023) # Yay on record per crash
```

### Mecanical Issues
The Factor data set deals with mechanical issues in a vehicle. This data set can record multiple problems per vehicle. Here I look at the frequency of the different combinations of accidents and then decide what to do with them. See Created Variable Dictionary for more details.

```{r}
# Factor has multiple entries per vehicle. Lets make it binary.
data_list[[4]] |>
  count(ST_CASE) # Illustrating issue

data_list[[4]] |>
  count(VEHICLECC) # not many instances of most

wideFac <- data_list[[4]] %>% # Pivoting data set to have unique columns and correct number of rows.
  mutate(VEHICLECC = if_else(VEHICLECC == 0, 101, VEHICLECC)) %>%
  pivot_wider(
    id_cols = c(ST_CASE, VEH_NO),
    names_from = VEHICLECC,
    values_from = VEHICLECC,
    values_fill = 0,
    values_fn = max
  )

uniteCountFac <- wideFac %>% # creating a data set of unique combos
  #mutate(flag = if_else(
    #rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(2, 3, 11, 12))) > 0, 1, 0
  #)) %>%
  unite("combo", -c(ST_CASE, VEH_NO), sep = "_") %>%
  count(combo, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100,
         cum_pct = cumsum(pct))

# THIS IS INCREDIBLY SPARCE! 98.07548% are unknown/non-noted

# Barbarize if there is an issue at all
data_list[[4]] <- data_list[[4]] |>
  mutate(MECISSUE = case_when(
    VEHICLECC == 0 ~ 0, # No mechanical Issue
    VEHICLECC %in% c(98, 99) ~ 2, # Unknown/not reported
    TRUE ~ 1 # Mechanical Issue
  )) %>%
  select(-VEHICLECC) %>%
  distinct()

data_list[[4]] %>%
  count(MECISSUE)

data_list[[4]] %>%
  count(ST_CASE) %>%
  arrange(desc(n))
```

### Distraction Issues
The distractions can have multiple per vehicle. Here I look at their unique combinations and transform this into one variable.

```{r}
# Multiple Distractions listed

data_list[[5]] |>
  count(ST_CASE)

data_list[[5]] |>
  count(DRDISTRACT)

wideDis <- data_list[[5]] %>% # Pivoting data set to have unique columns and correct number of rows.
  mutate(DRDISTRACT = if_else(DRDISTRACT == 0, 101, DRDISTRACT)) %>%
  pivot_wider(
    id_cols = c(ST_CASE, VEH_NO),
    names_from = DRDISTRACT,
    values_from = DRDISTRACT,
    values_fill = 0,
    values_fn = max
  )

uniteCountDIS <- wideDis %>% # creating a data set of unique combos
  #mutate(flag = if_else(
    #rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(2, 3, 11, 12))) > 0, 1, 0
  #)) %>%
  unite("combo", -c(ST_CASE, VEH_NO), sep = "_") %>%
  count(combo, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100,
         cum_pct = cumsum(pct))

data_list[[5]] <- data_list[[5]] |>
  mutate(DISTRACT = case_when(
    DRDISTRACT == 0 ~ 0, # No Distraction
    DRDISTRACT %in% c(16) ~ 2, # Driver not present/unknown if present
    DRDISTRACT %in% c(96,99) ~ 3, # Unknown/not reported
    TRUE ~ 1 # Driver distracted
  )) %>%
  distinct()

data_list[[5]] %>%
  count(ST_CASE) %>%
  arrange(desc(n))
```


### Vision Issue
```{r}
data_list[[7]] |>
  count(ST_CASE)

data_list[[7]] |>
  count(VISION)

wideVIS <- data_list[[7]] %>% # Pivoting data set to have unique columns and correct number of rows.
  mutate(VISION = if_else(VISION == 0, 101, VISION)) %>%
  pivot_wider(
    id_cols = c(ST_CASE, VEH_NO),
    names_from = VISION,
    values_from = VISION,
    values_fill = 0,
    values_fn = max
  )

uniteCountVIS<- wideVIS %>% # creating a data set of unique combos
  #mutate(flag = if_else(
    #rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(2, 3, 11, 12))) > 0, 1, 0
  #)) %>%
  unite("combo", -c(ST_CASE, VEH_NO), sep = "_") %>%
  count(combo, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100,
         cum_pct = cumsum(pct))

data_list[[7]] <- data_list[[7]] |>
  mutate(VIS = case_when(
    VISION == 0 ~ 0, # No
    VISION %in% c(95) ~ 2, #Driver not present/unknown if not present
    VISION %in% c(99) ~ 3, #Unknown
    TRUE ~ 1 # Vision obscured
  )) %>%
  distinct()

data_list[[7]] %>%
  count(ST_CASE) %>%
  arrange(desc(n))
```


### Checking all Vehicle variables Balpark same
```{r}
nrow(data_list[[4]])
nrow(data_list[[5]])
nrow(data_list[[7]])
```

## Joining Non-Vehicle Files
```{r}
crashes2023 <- left_join(data_list[[3]], data_list[[1]], by = join_by(ST_CASE))

crashes2023 %>% # One record per person??
  group_by(ST_CASE, PER_NO, VEH_NO) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n)) %>%
  count(n)

nrow(crashes2023) ## Checking the Number of rows didn't increase
nrow(person2023)

crashes2023 <- left_join(crashes2023, wideNM, by = join_by(ST_CASE, PER_NO, VEH_NO))

nrow(crashes2023) ## Checking the Number of rows didn't increase
nrow(person2023)

crashes2023 %>% # One record per person??
  group_by(ST_CASE, PER_NO, VEH_NO) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n)) %>%
  count(n)
```

## Joining Vehicles and Vehicle Level Data

```{r}
nrow(crashInishator)
# Join other vehicle factors to sparce vehicle matrix
crashInishator <- left_join(crashInishator, data_list[[4]], by = join_by('ST_CASE', 'VEH_NO'))
nrow(crashInishator)
crashInishator1 <- left_join(crashInishator, data_list[[5]], by = join_by('ST_CASE', 'VEH_NO'))
nrow(crashInishator)
crashInishator2 <- left_join(crashInishator, data_list[[7]], by = join_by('ST_CASE', 'VEH_NO'))
nrow(crashInishator)
```

## Join Vehicle Data to Non-Vehicle Level
```{r}
crashes2023test <- left_join(crashes2023, crashInishator, by = join_by('ST_CASE', 'VEH_NO'))
```

