---
title: "566 Project Data Exploration"
author: "North Carpenter"
date: "2026-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(sf)
library(ggplot2)
library(lubridate)
```

# Importing Data

```{r}
# This block will import key data. It is imparative that you structure your
# files in the same way or this will not work. Set up this file in the 
# Y-crash-project folder somewhere on your computer and at the same level of the 
# project folder put the 566 project data folder.
#
# You should be able to import the files in 566 in this chunk only without issue.

# Importing Data

# Import any csv's here
accident2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/accident.csv")
person2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/PERSON.csv")
vehicle2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/VEHICLE.csv")
drdistract2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/distract.csv")
factor2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/factor.csv")
nmCrash2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/nmcrash.csv")
cEvent2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/cevent.csv")
vission2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/vision.csv")

# Just Curios
drugs2023 <- read.csv("../566 Project Data/FARS2023NationalCSV/drugs.csv")
```

# Cleaning
## Setting Up Variables to Keep
```{r}
merging_Vars <- c("ST_CASE", "VEH_NO", "PER_NO")
accident_keep <- c("VE_FORMS", "MONTH", "DAY", "YEAR", "HOUR", "MINUTE", "DAY_WEEK", "ROUTE", "RUR_URB", "RELJCT1", "LGT_COND", "WEATHER", "ARR_HOUR", "ARR_MIN")
vehicle_keep <- c("TRAV_SP", "DR_DRINK", "PREV_ACC", "VSPD_LIM", "VPAVETYP", "VSURCOND", "P_CRASH1", "M_HARM")
person_keep <- c("AGE", "SEX", "PER_TYP", "INJ_SEV", "DRUGS", "DEATH_MO", "DEATH_DA", "DEATH_YR", "DEATH_TM", "DEATH_HR", "DEATH_MN", "LAG_HRS", "LAG_MINS") #Record Sex of Driver Only, fix type, fix type, complex death after indicator
vision_keep <- c("VISION")
fact_keep <- c("VEHICLECC")
dist_keep <- c("DRDISTRACT")
NMCC_keep <- c("NMCC")
```

## Triming Files
```{r}
data_list <- list(accident2023, vehicle2023, person2023, factor2023, drdistract2023, nmCrash2023, vission2023)
var_list <- list(accident_keep, vehicle_keep, person_keep, fact_keep, dist_keep, NMCC_keep, vision_keep)

data_list <- Map(function(df, vars){
  df %>% select(any_of(c(merging_Vars, vars)))
}, data_list, var_list)
```

## Check Worked
```{r}
lapply(data_list, head)
```

## Creating Variables For Joins
Dealing with necessary variable manipulation prior to joining. Some variables have multiple entries for the unit they represent. Here I investigate that and transform the variables so the sets can be merged. See "Created Variable Dictionary" Google doc for a condensed description of what is done below.
### NMCrash
```{r}
data_list[[6]] %>% # There are multiple entries per combo of case, person, vehicle
  group_by(ST_CASE, PER_NO, VEH_NO) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n)) %>%
  count(n)

data_list[[6]] %>% # base frequency in unpivoted dataset
  count(NMCC)

wideNM <- data_list[[6]] %>% # Pivoting data set to have unique columns and correct number of rows.
  mutate(NMCC = if_else(NMCC == 0, 101, NMCC)) %>%
  pivot_wider(
    id_cols = c(ST_CASE, PER_NO, VEH_NO),
    names_from = NMCC,
    values_from = NMCC,
    values_fill = 0,
    values_fn = max
  )

uniteCount <- wideNM %>% # creating a data set of unique combos
  mutate(flag = if_else(
    rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(2, 3, 11, 12))) > 0, 1, 0
  )) %>%
  unite("combo", -c(ST_CASE, PER_NO, VEH_NO, flag), sep = "_") %>%
  count(combo, flag, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100,
         cum_pct = cumsum(pct))

uniteCount %>%
  filter(flag == 1) %>%
  summarise(total_pct = sum(pct))

# DEFFINITIONS OF NEW VARIABLES
wideNM <- wideNM %>%
  mutate(
    NMCCSimple = case_when(
      rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(101))) > 0 ~ 0,
      rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(98, 99))) > 0 ~ 5,
      rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(2))) > 0 ~ 1,
      rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(19))) > 0 ~ 3,
      rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(2, 3, 4, 11, 12))) > 0 ~ 3,
      TRUE ~ 4
    )
  )

# Drop Unused columns
wideNM <- wideNM %>%
  select(c(ST_CASE, PER_NO, VEH_NO, NMCCSimple))
```

### Initiating Vehicle Identification
Making a list of the vehicle that initiate each accident.
```{r}
# Get vehicle number for first vehicle to act in the event order
acLVLcEvent <- pivot_wider(cEvent2023, names_from = EVENTNUM, values_from = VNUMBER1) %>%
  filter(!is.na(`1`)) %>%
  rename('vEvent1' = '1') %>%
  subset(select = c('ST_CASE', 'vEvent1'))
head(acLVLcEvent)

# Dropping all rows not the initiating vehicle
crashInishator <- left_join(data_list[[2]], acLVLcEvent, by = 'ST_CASE') %>%
   filter(vEvent1 == VEH_NO) %>% # Keep only first vehicles
  select(-vEvent1) # Drop indicator now equal to VEH_NO
head(crashInishator)

nrow(crashInishator)
nrow(accident2023) # Yay on record per crash
```

### Mecanical Issues
The Factor data set deals with mechanical issues in a vehicle. This data set can record multiple problems per vehicle. Here I look at the frequency of the different combinations of accidents and then decide what to do with them. See Created Variable Dictionary for more details.

```{r}
# Factor has multiple entries per vehicle. Lets make it binary.
data_list[[4]] |>
  count(ST_CASE) # Illustrating issue

data_list[[4]] |>
  count(VEHICLECC) # not many instances of most

wideFac <- data_list[[4]] %>% # Pivoting data set to have unique columns and correct number of rows.
  mutate(VEHICLECC = if_else(VEHICLECC == 0, 101, VEHICLECC)) %>%
  pivot_wider(
    id_cols = c(ST_CASE, VEH_NO),
    names_from = VEHICLECC,
    values_from = VEHICLECC,
    values_fill = 0,
    values_fn = max
  )

uniteCountFac <- wideFac %>% # creating a data set of unique combos
  #mutate(flag = if_else(
    #rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(2, 3, 11, 12))) > 0, 1, 0
  #)) %>%
  unite("combo", -c(ST_CASE, VEH_NO), sep = "_") %>%
  count(combo, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100,
         cum_pct = cumsum(pct))

# THIS IS INCREDIBLY SPARCE! 98.07548% are unknown/non-noted

# Barbarize if there is an issue at all
wideFac <- wideFac %>%
  mutate(
    MECISSUE = case_when(
      rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(101))) > 0 ~ 0,
      rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(98, 99))) > 0 ~ 2,
      TRUE ~ 1
    )
  )

wideFac %>%
  count(MECISSUE)

wideFac %>%
  count(ST_CASE) %>%
  arrange(desc(n))

wideFac <- wideFac %>%
  select(c(ST_CASE, VEH_NO, MECISSUE))
```

### Distraction Issues
The distractions can have multiple per vehicle. Here I look at their unique combinations and transform this into one variable.

```{r}
# Multiple Distractions listed

data_list[[5]] |>
  count(ST_CASE)

data_list[[5]] |>
  count(DRDISTRACT)

wideDis <- data_list[[5]] %>% # Pivoting data set to have unique columns and correct number of rows.
  mutate(DRDISTRACT = if_else(DRDISTRACT == 0, 101, DRDISTRACT)) %>%
  pivot_wider(
    id_cols = c(ST_CASE, VEH_NO),
    names_from = DRDISTRACT,
    values_from = DRDISTRACT,
    values_fill = 0,
    values_fn = max
  )

uniteCountDIS <- wideDis %>% # creating a data set of unique combos
  #mutate(flag = if_else(
    #rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(2, 3, 11, 12))) > 0, 1, 0
  #)) %>%
  unite("combo", -c(ST_CASE, VEH_NO), sep = "_") %>%
  count(combo, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100,
         cum_pct = cumsum(pct))

wideDis <- wideDis %>%
  mutate(
    DISTRACT = case_when(
      rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(101))) > 0 ~ 0,
      rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(16))) > 0 ~ 2,
      rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(96, 99))) > 0 ~ 3,
      TRUE ~ 1
    )
  )

wideDis %>%
  count(ST_CASE) %>%
  arrange(desc(n))

wideDis <- wideDis %>%
  select(c(ST_CASE, VEH_NO, DISTRACT))
```


### Vision Issue
```{r}
data_list[[7]] |>
  count(ST_CASE)

data_list[[7]] |>
  count(VISION)

wideVIS <- data_list[[7]] %>% # Pivoting data set to have unique columns and correct number of rows.
  mutate(VISION = if_else(VISION == 0, 101, VISION)) %>%
  pivot_wider(
    id_cols = c(ST_CASE, VEH_NO),
    names_from = VISION,
    values_from = VISION,
    values_fill = 0,
    values_fn = max
  )

uniteCountVIS<- wideVIS %>% # creating a data set of unique combos
  #mutate(flag = if_else(
    #rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(2, 3, 11, 12))) > 0, 1, 0
  #)) %>%
  unite("combo", -c(ST_CASE, VEH_NO), sep = "_") %>%
  count(combo, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100,
         cum_pct = cumsum(pct))

wideVIS <- wideVIS %>%
  mutate(
    VIS = case_when(
      rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(101))) > 0 ~ 0,
      rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(95))) > 0 ~ 2,
      rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(99))) > 0 ~ 3,
      TRUE ~ 1
    )
  )

wideVIS %>%
  count(ST_CASE) %>%
  arrange(desc(n))

wideVIS <- wideVIS %>%
  select(c(ST_CASE, VEH_NO, VIS))
```

### Driver Characeristics
```{r}
driverData <- data_list[[3]] %>% 
  select(c(ST_CASE, PER_NO, VEH_NO, PER_TYP, AGE, SEX, DRUGS)) %>% 
  filter(PER_TYP == 1) %>%
  rename(DR_AGE = AGE, DR_SEX = SEX, DR_DRUGS = DRUGS) %>%
  select(-c(PER_NO, PER_TYP))
head(driverData)
```



## Joining Non-Vehicle Files
```{r}
crashes2023 <- left_join(data_list[[3]], data_list[[1]], by = join_by(ST_CASE))

crashes2023 %>% # One record per person??
  group_by(ST_CASE, PER_NO, VEH_NO) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n)) %>%
  count(n)

nrow(crashes2023) ## Checking the Number of rows didn't increase
nrow(person2023)

crashes2023 <- left_join(crashes2023, wideNM, by = join_by(ST_CASE, PER_NO, VEH_NO))

nrow(crashes2023) ## Checking the Number of rows didn't increase
nrow(person2023)

crashes2023 %>% # One record per person??
  group_by(ST_CASE, PER_NO, VEH_NO) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n)) %>%
  count(n)
```

## Joining Vehicles and Vehicle Level Data

```{r}
nrow(crashInishator)
# Join other vehicle factors to sparce vehicle matrix
crashInishator <- left_join(crashInishator, wideFac, by = join_by('ST_CASE', 'VEH_NO'))
nrow(crashInishator)
crashInishator <- left_join(crashInishator, wideDis, by = join_by('ST_CASE', 'VEH_NO'))
nrow(crashInishator)
crashInishator <- left_join(crashInishator, wideVIS, by = join_by('ST_CASE', 'VEH_NO'))
nrow(crashInishator)
crashInishator <- left_join(crashInishator, driverData, by = join_by('ST_CASE', 'VEH_NO'))
nrow(crashInishator)

# Cleaning up columns
crashInishator <- rename(crashInishator, CrashVehicle = VEH_NO) # make initiation vehicle distinct
head(crashInishator)
```

## Join Vehicle Data to Non-Vehicle Level
```{r}
nrow(crashes2023)
crashes2023 <- left_join(crashes2023, crashInishator, by = join_by('ST_CASE'))
nrow(crashes2023)
nrow(person2023)
head(crashes2023)
```



# New Variables
## PedDeath
If this entry represents a pedestrian death. See Created Variable Dictionary.
```{r}
crashes2023 <- crashes2023 %>%
  mutate(PED_DEATH = if_else(PER_TYP %in% c(5, 8) & INJ_SEV == 4, 1, 0))
head(crashes2023)

sum(crashes2023$PED_DEATH)
```

## Emergency Vehicle Arrival and Death
Time in minutes until emergency services arrive. See Created Variable Dictionary. Vibe coded and read through.
```{r}
# Compute EMSAD (EMS Arrival Difference) variable
# Goal: categorize whether EMS arrival time matters at the person level
#
# IMPORTANT CONTEXT: Deaths can be recorded up to 30 days after the crash.
# We must use full datetime (year, month, day, hour, minute) for the
# crash-to-death comparison, not just clock time.
#
# Algorithm:
#   1. If death occurred within 5 minutes of crash → EMSAD = 0
#      (EMS arrival irrelevant; person died too quickly for EMS to matter)
#   2. Otherwise → EMSAD = minutes elapsed from crash to EMS arrival
#      (EMS arrival is still assumed within 24 hours; no date is recorded for it)

crashes2023 <- crashes2023 %>%
  mutate(
    # --- Step 1: Build a full datetime for the crash ---
    # Uses lubridate::make_datetime() to combine the separate date/time
    # columns into a single POSIXct timestamp for accurate differencing.
    # Adjust column name prefixes to match your actual crash date columns.
    crash_datetime = make_datetime(
      year  = YEAR,
      month = MONTH,
      day   = DAY,
      hour  = HOUR,
      min   = MINUTE
    ),

    # --- Step 2: Build a full datetime for the death ---
    # Death can occur up to 30 days after the crash, so we must include
    # the death's own YEAR, MONTH, and DAY — not just the time of day.
    death_datetime = make_datetime(
      year  = DEATH_YR,
      month = DEATH_MO,
      day   = DEATH_DA,
      hour  = DEATH_HR,
      min   = DEATH_MN
    ),

    # --- Step 3: Compute crash-to-death difference in minutes ---
    # as.numeric(..., units = "mins") gives a plain numeric value,
    # which is easier to threshold (e.g., <= 5) than a difftime object.
    crash_to_death_min = as.numeric(
      difftime(death_datetime, crash_datetime, units = "mins")
    ),

    # --- Step 4: Convert EMS arrival time to total minutes since midnight ---
    # EMS arrival has no date, so we use the same midnight-crossing trick
    # as before: if the arrival clock time is earlier than the crash clock
    # time, EMS arrived after midnight → add 1440 minutes.
    # This is valid because EMS arrival is assumed within 24 hours.
    crash_total_min = HOUR * 60 + MINUTE,
    ems_total_min   = ARR_HOUR * 60 + ARR_MIN,

    crash_to_ems_min = if_else(
      ems_total_min >= crash_total_min,
      ems_total_min - crash_total_min,
      ems_total_min - crash_total_min + 1440  # EMS arrived after midnight
    ),

    # --- Step 5: Apply the EMSAD algorithm ---
    # Rule 1: If the person died within 5 minutes of the crash (regardless
    #         of what day, since difftime handles the full date span),
    #         EMS arrival time is immaterial → assign 0.
    # Rule 2: Otherwise, record how many minutes after the crash EMS arrived,
    #         so downstream models can assess whether faster response
    #         could have changed the outcome.
    EMSAD = if_else(
      crash_to_death_min <= 5,
      0,                  # Died too soon; EMS arrival irrelevant
      crash_to_ems_min    # Minutes from crash until EMS arrived on scene
    )
  ) %>%

  # --- Step 6: Drop all intermediate columns used for computation ---
  select(-crash_datetime, -death_datetime,
         -crash_to_death_min, -crash_total_min,
         -ems_total_min, -crash_to_ems_min)
```

# Dropping Unused Variables
```{r}
head(crashes2023)
ncol(crashes2023)

# Dropping Predictors of outcome
crashes2023 <- crashes2023 %>%
  select(-PER_TYP, -INJ_SEV)

# Dropping Variables used to construct other variables but we don't want
crashes2023 <- crashes2023 %>%
  select(-YEAR, -MONTH, -DAY, -HOUR, -MINUTE, -DEATH_YR, -DEATH_MO, -DEATH_DA, -DEATH_HR, -DEATH_MN, DEATH_TM, -ARR_HOUR, -ARR_MIN, LAG_HRS, LAG_MINS)

# Dropping Non-predictive variables for the victm
crashes2023 <- crashes2023 %>%
  select(-SEX)

ncol(crashes2023)
```

# Export 2023
Save Data for 2023
```{r}
write.csv(crashes2023, "../566 Project Data/crashes2023.csv")
```

# Drop Data
```{r}
rm(list = setdiff(
  ls()[sapply(ls(), function(x) is.data.frame(get(x)))],  # all dfs in env
  c("crashes2023")                   # except these
))
```

# Functions for all transformations and Merges.

## Importing Data
```{r}
importYears <- function(start, end, files, fileLoc)
{
  # Imports CSVs for a range of years into a named list
  # ====================================================
  # Inputs
  # - start:   start year of data as integer
  # - end:     end year of data as integer
  # - files:   character vector of CSV filenames (e.g. "accident.csv")
  # - fileLoc: path to the project data folder
  # Outputs
  # - named list of data frames, keyed as "YYYY_filename.csv"
  years <- start:end
  combos <- expand.grid(year = years, file = files, stringsAsFactors = FALSE)
  
  paths <- paste0(fileLoc, "/FARS", combos$year, "NationalCSV/", combos$file)
  crashFiles <- lapply(paths, read.csv)
  
  # Preserve Names
  names(crashFiles) <- paste0(combos$year, "_", combos$file)
  
  return(crashFiles)
}

# Files for Variables currently used
files <- c("accident.csv", "PERSON.csv", "VEHICLE.csv", "distract.csv", "factor.csv", "nmcrash.csv", "cevent.csv", "vision.csv")
fileLoc <- "../566 Project Data"

FARS2023 <- importYears(2023, 2023, files, fileLoc)
```

## Setting Up Variables to Keep
```{r}
merging_Vars <- c("ST_CASE", "VEH_NO", "PER_NO")
accident_keep <- c("VE_FORMS", "MONTH", "DAY", "YEAR", "HOUR", "MINUTE",
                   "DAY_WEEK", "ROUTE", "RUR_URB", "RELJCT1", "LGT_COND",
                   "WEATHER", "ARR_HOUR", "ARR_MIN")
vehicle_keep <- c("TRAV_SP", "DR_DRINK", "PREV_ACC", "VSPD_LIM", "VPAVETYP",
                  "VSURCOND", "P_CRASH1", "M_HARM")
person_keep <- c("AGE", "SEX", "PER_TYP", "INJ_SEV", "DRUGS", "DEATH_MO",
                 "DEATH_DA", "DEATH_YR", "DEATH_TM", "DEATH_HR", "DEATH_MN",
                 "LAG_HRS", "LAG_MINS")
vision_keep <- c("VISION")
fact_keep <- c("VEHICLECC")
dist_keep <- c("DRDISTRACT")
NMCC_keep <- c("NMCC")
```

## Triming Files
```{r}
# Map file names to their keep variables
var_map <- list(
  "accident.csv" = accident_keep,
  "VEHICLE.csv"  = vehicle_keep,
  "PERSON.csv"   = person_keep,
  "factor.csv"   = fact_keep,
  "distract.csv" = dist_keep,
  "nmcrash.csv"  = NMCC_keep,
  "vision.csv"   = vision_keep
)

trimYear <- function(crash_data, year, var_map) {
  # Trims a single year of data to only the columns needed for analysis
  # ====================================================================
  # Inputs
  # - crash_data: named list of data frames from importYears
  # - year:       integer year to trim
  # - var_map:    named list mapping filenames to their keep variables
  # Outputs
  # - named list of trimmed data frames for the given year, keyed by filename
  Map(function(file, vars) {
    key <- paste0(year, "_", file)
    crash_data[[key]] %>% select(any_of(c(merging_vars, vars)))
  }, names(var_map), var_map)
}


trimAllYears <- function(crash_data, start, end, var_map) {
  # Trims all years of data to only the columns needed for analysis
  # ================================================================
  # Inputs
  # - crash_data: named list of data frames from importYears
  # - start:      start year of data as integer
  # - end:        end year of data as integer
  # - var_map:    named list mapping filenames to their keep variables
  # Outputs
  # - nested named list keyed as result[["YYYY"]][["filename.csv"]]
  years <- start:end
  result <- lapply(years, function(year) trimYear(crash_data, year, var_map))
  names(result) <- years
  return(result)
}
```

## Variable Creation

### Person Level Data
#### NMCCSimple

### Vehicle Level Data
#### MECISSUE
#### DISTRACT
#### VIS

## Joins

## Full Data Variable Creation
### PedDeath
### EMSAD
##
