---
title: "anyYearData"
author: "North C"
date: "2026-05-28"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(sf)
library(ggplot2)
library(lubridate)
library(mice)
library(sf)
```

# Functions for all transformations and Merges.

## Importing Data
```{r}
importYears <- function(start, end, files, fileLoc)
{
  # Imports CSVs for a range of years into a named list
  # ====================================================
  # Inputs
  # - start:   start year of data as integer
  # - end:     end year of data as integer
  # - files:   character vector of CSV filenames (e.g. "accident.csv")
  # - fileLoc: path to the project data folder
  # Outputs
  # - named list of data frames, keyed as "YYYY_filename.csv"
  years <- start:end
  combos <- expand.grid(year = years, file = files, stringsAsFactors = FALSE)
  
  paths <- paste0(fileLoc, "/FARS", combos$year, "NationalCSV/", combos$file)
  crashFiles <- lapply(paths, function(p) as.data.frame(data.table::fread(p)))
  
  # Preserve Names
  names(crashFiles) <- paste0(combos$year, "_", combos$file)
  
  return(crashFiles)
}
```

## Triming Files
```{r}
trimYear <- function(crash_data, year, var_map, merging_Vars) {
  # Trims a single year of data to only the columns needed for analysis
  # ====================================================================
  # Inputs
  # - crash_data: named list of data frames from importYears
  # - year:       integer year to trim
  # - var_map:    named list mapping filenames to their keep variables
  # Outputs
  # - named list of trimmed data frames for the given year, keyed by filename
  Map(function(file, vars) {
    key <- paste0(year, "_", file)
    crash_data[[key]] %>% select(any_of(c(merging_Vars, vars)))
  }, names(var_map), var_map)
}


trimAllYears <- function(crash_data, start, end, var_map, merging_Vars) {
  # Trims all years of data to only the columns needed for analysis
  # ================================================================
  # Inputs
  # - crash_data: named list of data frames from importYears
  # - start:      start year of data as integer
  # - end:        end year of data as integer
  # - var_map:    named list mapping filenames to their keep variables
  # Outputs
  # - nested named list keyed as result[["YYYY"]][["filename.csv"]]
  years <- start:end
  result <- lapply(years, function(year) trimYear(crash_data, year, var_map, merging_Vars))
  names(result) <- years
  return(result)
}
```

## Variable Creation

### Person Level Data
#### NMCCSimple
```{r}
processNMCrash <- function(nm_df) {
  # Pivots and simplifies nmcrash data into a single categorical column
  # ====================================================================
  # Inputs
  # - nm_df: a single nmcrash data frame with columns ST_CASE, PER_NO, VEH_NO, NMCC
  # Outputs
  # - trimmed data frame with ST_CASE, PER_NO, VEH_NO, and NMCCSimple
  
  # Normalize legacy column name to NMCC before processing
  if ("MTM_CRSH" %in% names(nm_df) && !"NMCC" %in% names(nm_df)) {
    nm_df <- rename(nm_df, NMCC = MTM_CRSH)
  } else if (!"NMCC" %in% names(nm_df)) {
    stop("Input data frame must contain either 'NMCC' or 'MTM_CRSH' column.")
  }
  
  nm_df %>%
    mutate(NMCC = if_else(NMCC == 0, 101, NMCC)) %>%
    pivot_wider(
      id_cols = c(ST_CASE, PER_NO, VEH_NO),
      names_from = NMCC,
      values_from = NMCC,
      values_fill = 0,
      values_fn = max
    ) %>%
    mutate(
      NMCCSimple = case_when(
        rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(101))) > 0 ~ 0,
        rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(98, 99))) > 0 ~ 5,
        rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(2))) > 0 ~ 1,
        rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(19))) > 0 ~ 3,
        rowSums(sapply(select(., -c(ST_CASE, PER_NO, VEH_NO)), `%in%`, c(2, 3, 4, 11, 12))) > 0 ~ 3,
        TRUE ~ 4
      )
    ) %>%
    select(ST_CASE, PER_NO, VEH_NO, NMCCSimple)
}
```

#### Driver Data
```{r}
processDriverData <- function(person_df) {
  # Filters and renames person data to retain only driver-level variables
  # ======================================================================
  # Inputs
  # - person_df: a single person data frame with columns ST_CASE, PER_NO,
  #              VEH_NO, PER_TYP, AGE, SEX, DRUGS
  # Outputs
  # - data frame filtered to drivers only with renamed demographic columns

  person_df %>%
    filter(PER_TYP == 1) %>%
    rename(DR_AGE = AGE, DR_SEX = SEX, DR_DRUGS = DRUGS) %>%
    select(ST_CASE, VEH_NO, DR_AGE, DR_SEX, DR_DRUGS)
}
```

### Vehicle Level Data
Identifying First Vehicle in the Crash
```{r}
processInitiatingVehicle <- function(cevent_df, vehicle_df) {
  # Identifies and isolates the initiating vehicle for each crash event
  # ===================================================================
  # Inputs
  # - cevent_df:  a single cevent data frame with columns ST_CASE, EVENTNUM, VNUMBER1
  # - vehicle_df: a single vehicle data frame with columns ST_CASE, VEH_NO
  # Outputs
  # - vehicle data frame filtered to only the first vehicle in each crash event

  first_vehicle <- cevent_df %>%
    pivot_wider(names_from = EVENTNUM, values_from = VNUMBER1) %>%
    filter(!is.na(`1`)) %>%
    rename(vEvent1 = `1`) %>%
    select(ST_CASE, vEvent1)

  vehicle_df %>%
    left_join(first_vehicle, by = "ST_CASE") %>%
    filter(vEvent1 == VEH_NO) %>%
    select(-vEvent1)
}
```

#### MECISSUE
```{r}
processMechanicalIssue <- function(factor_df) {
  # Pivots and simplifies vehicle factor data into a single mechanical issue indicator
  # ==================================================================================
  # Inputs
  # - factor_df: a single factor data frame with columns ST_CASE, VEH_NO, VEHICLECC
  # Outputs
  # - trimmed data frame with ST_CASE, VEH_NO, and MECISSUE
  
  if ("MFACTOR" %in% names(factor_df) && !"VEHICLECC" %in% names(factor_df)) {
    factor_df <- rename(factor_df, VEHICLECC = MFACTOR)
  } else if (!"VEHICLECC" %in% names(factor_df)) {
    stop("Input data frame must contain either 'MFACTOR' or 'VEHICLECC' column.")
  }

  factor_df %>%
    mutate(VEHICLECC = if_else(VEHICLECC == 0, 101, VEHICLECC)) %>%
    pivot_wider(
      id_cols = c(ST_CASE, VEH_NO),
      names_from = VEHICLECC,
      values_from = VEHICLECC,
      values_fill = 0,
      values_fn = max
    ) %>%
    mutate(
      MECISSUE = case_when(
        rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(101))) > 0 ~ 0,
        rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(98, 99))) > 0 ~ 2,
        TRUE ~ 1
      )
    ) %>%
    select(ST_CASE, VEH_NO, MECISSUE)
}
```

#### DISTRACT
```{r}
processDistraction <- function(distract_df) {
  # Pivots and simplifies driver distraction data into a single distraction indicator
  # ==================================================================================
  # Inputs
  # - distract_df: a single distract data frame with columns ST_CASE, VEH_NO, DRDISTRACT
  # Outputs
  # - trimmed data frame with ST_CASE, VEH_NO, and DISTRACT
  
  if ("MDRDSTRD" %in% names(distract_df) && !"DRDISTRACT" %in% names(distract_df)) {
    distract_df <- rename(distract_df, DRDISTRACT = MDRDSTRD)
  } else if (!"DRDISTRACT" %in% names(distract_df)) {
    stop("Input data frame must contain either 'MDRDSTRD' or 'DRDISTRACT' column.")
  }

  distract_df %>%
    mutate(DRDISTRACT = if_else(DRDISTRACT == 0, 101, DRDISTRACT)) %>%
    pivot_wider(
      id_cols = c(ST_CASE, VEH_NO),
      names_from = DRDISTRACT,
      values_from = DRDISTRACT,
      values_fill = 0,
      values_fn = max
    ) %>%
    mutate(
      DISTRACT = case_when(
        rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(101))) > 0 ~ 0,
        rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(16))) > 0 ~ 2,
        rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(96, 99))) > 0 ~ 3,
        TRUE ~ 1
      )
    ) %>%
    select(ST_CASE, VEH_NO, DISTRACT)
}
```

#### VIS
```{r}
processVision <- function(vision_df) {
  # Pivots and simplifies driver vision data into a single vision obstruction indicator
  # ====================================================================================
  # Inputs
  # - vision_df: a single vision data frame with columns ST_CASE, VEH_NO, VISION
  # Outputs
  # - trimmed data frame with ST_CASE, VEH_NO, and VIS
  
  if ("MVISOBSC" %in% names(vision_df) && !"VISION" %in% names(vision_df)) {
    vision_df <- rename(vision_df, VISION = MVISOBSC)
  } else if (!"VISION" %in% names(vision_df)) {
    stop("Input data frame must contain either 'MVISOBSC' or 'VISION' column.")
  }

  vision_df %>%
    mutate(VISION = if_else(VISION == 0, 101, VISION)) %>%
    pivot_wider(
      id_cols = c(ST_CASE, VEH_NO),
      names_from = VISION,
      values_from = VISION,
      values_fill = 0,
      values_fn = max
    ) %>%
    mutate(
      VIS = case_when(
        rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(101))) > 0 ~ 0,
        rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(95))) > 0 ~ 2,
        rowSums(sapply(select(., -c(ST_CASE, VEH_NO)), `%in%`, c(99))) > 0 ~ 3,
        TRUE ~ 1
      )
    ) %>%
    select(ST_CASE, VEH_NO, VIS)
}
```

## Joins + Prossessing
```{r}
mergeYear <- function(df_list, personLVLName = "personLVL.csv") {
  # Processes and joins all files into a single person level data frame for one year
  # =================================================================================
  # Inputs
  # - df_list:       named list of raw trimmed data frames for one year
  # - personLVLName: name to save the intermediate person level file under
  # Outputs
  # - single flat person level data frame with initiating driver characteristics

  # Step 1: process all file level transformations
  df_list[["nmcrash.csv"]]  <- processNMCrash(df_list[["nmcrash.csv"]])
  df_list[["factor.csv"]]   <- processMechanicalIssue(df_list[["factor.csv"]])
  df_list[["distract.csv"]] <- processDistraction(df_list[["distract.csv"]])
  df_list[["vision.csv"]]   <- processVision(df_list[["vision.csv"]])
  df_list[["VEHICLE.csv"]]  <- processInitiatingVehicle(df_list[["cevent.csv"]], df_list[["VEHICLE.csv"]])

  # Step 2: build and save person level file into the list
  df_list[[personLVLName]] <- df_list[["PERSON.csv"]] %>%
    left_join(df_list[["accident.csv"]], by = "ST_CASE") %>%
    left_join(df_list[["nmcrash.csv"]],  by = c("ST_CASE", "VEH_NO", "PER_NO"))

  # Step 3: build driver level file from the person level file
  df_list[["driverLVL.csv"]] <- processDriverData(df_list[[personLVLName]])

  # Step 4: join initiating vehicle and driver characteristics to each person
  initiating_vehicle <- df_list[["VEHICLE.csv"]] %>%
    left_join(df_list[["factor.csv"]],    by = c("ST_CASE", "VEH_NO")) %>%
    left_join(df_list[["distract.csv"]], by = c("ST_CASE", "VEH_NO")) %>%
    left_join(df_list[["vision.csv"]],   by = c("ST_CASE", "VEH_NO")) %>%
    left_join(df_list[["driverLVL.csv"]], by = c("ST_CASE", "VEH_NO")) %>%
    rename(CrashVehicle = VEH_NO)

  df_list[[personLVLName]] %>%
    left_join(initiating_vehicle,        by = "ST_CASE")
}
```

## Full Data Variable Creation
### RUL_URB, PED_DEATH, EMSAD
```{r}
addDerivedVariables <- function(crashes_df) {
  # Adds person and crash level derived variables to the full dataset
  # =================================================================
  # Inputs
  # - crashes_df: full joined data frame with columns PER_TYP, INJ_SEV,
  #               YEAR, MONTH, DAY, HOUR, MINUTE, DEATH_YR, DEATH_MO,
  #               DEATH_DA, DEATH_HR, DEATH_MN, ARR_HOUR, ARR_MIN
  # Outputs
  # - data frame with two additional columns:
  #   - PED_DEATH: 1 if row is a pedestrian death, 0 otherwise
  #   - EMSAD:     -1 if crash, death, or EMS arrival time is unknown,
  #                    canceled, or not applicable,
  #                 0 if person died within 5 min of crash,
  #                >0 minutes from crash to EMS arrival
  
  if (!"ROAD_FNC" %in% names(crashes_df)) {
    crashes_df$ROAD_FNC <- NA_integer_
  }

  crashes_df %>%
    mutate(
      
      # --- RUL_URB Handeling
      
      RUR_URB_Fix = case_when(
        YEAR > 2014 ~ RUR_URB,
        ROAD_FNC %in% c(1, 2, 3, 4, 5, 6, 9) ~ 1,
        ROAD_FNC %in% c(11, 12, 13, 14, 15, 16, 19) ~ 2,
        TRUE ~ 9
      ),
      
      # --- PED_DEATH ---
      PED_DEATH = if_else(PER_TYP %in% c(5, 8) & INJ_SEV == 4, 1, 0),

      # --- EMSAD: flag uncomputable cases ---
      # Crash time: HOUR 99 = unknown, MINUTE 99 = unknown, DAY 99 = unknown
      crash_time_unknown = HOUR == 99 | MINUTE == 99 | DAY == 99,
      # EMS time: ARR_HOUR 88 = unknown/canceled/not arrived
      #           ARR_MIN 99 = unknown, 98 = unknown if arrived,
      #                    97 = canceled, 88 = not applicable/notified
      ems_time_unknown = ARR_HOUR == 88 |
                         ARR_MIN %in% c(99, 98, 97, 88),
      # Death time: DEATH_YR 9999 = unknown, DEATH_HR 88 = not applicable,
      #             DEATH_HR 99 = unknown, DEATH_MN 88 = not applicable,
      #             DEATH_MN 99 = unknown
      death_time_unknown = DEATH_YR == 9999 |
                           DEATH_HR %in% c(88, 99) |
                           DEATH_MN %in% c(88, 99),

      # --- EMSAD: build intermediate datetimes ---
      # Substitute 0 for unknown times so make_datetime doesn't produce NA;
      # these rows will be overwritten with -1 in the final EMSAD step
      crash_datetime = make_datetime(
        year  = YEAR,
        month = MONTH,
        day   = DAY,
        hour  = if_else(crash_time_unknown, 0L, HOUR),
        min   = if_else(crash_time_unknown, 0L, MINUTE)
      ),
      death_datetime = make_datetime(
        year  = if_else(death_time_unknown, 0L, DEATH_YR),
        month = if_else(death_time_unknown, 1L, DEATH_MO),
        day   = if_else(death_time_unknown, 1L, DEATH_DA),
        hour  = if_else(death_time_unknown, 0L, DEATH_HR),
        min   = if_else(death_time_unknown, 0L, DEATH_MN)
      ),
      crash_to_death_min = as.numeric(
        difftime(death_datetime, crash_datetime, units = "mins")
      ),
      crash_total_min  = HOUR * 60 + MINUTE,
      ems_total_min    = ARR_HOUR * 60 + ARR_MIN,
      crash_to_ems_min = if_else(
        ems_total_min >= crash_total_min,
        ems_total_min - crash_total_min,
        ems_total_min - crash_total_min + 1440
      ),

      # --- EMSAD: final variable ---
      # -1 = crash, death, or EMS time unknown, canceled, or not applicable
      #  0 = died within 5 min of crash, EMS arrival irrelevant
      # >0 = minutes from crash to EMS arrival
      EMSAD = case_when(
        crash_time_unknown | ems_time_unknown | death_time_unknown ~ -1,
        crash_to_death_min <= 5                                    ~  0,
        TRUE                                                       ~ crash_to_ems_min
      )
    ) %>%
    select(-crash_datetime, -death_datetime,
           -crash_time_unknown, -ems_time_unknown, -death_time_unknown,
           -crash_to_death_min, -crash_total_min,
           -ems_total_min, -crash_to_ems_min, -RUR_URB, -ROAD_FNC) %>%
    rename(RUR_URB = RUR_URB_Fix)
}
```

### Data Coarsening
```{r}
coarsen <- function(crashes_df)
{
  # 6 categories
  crashes_df$M_HARM <- dplyr::case_when(
    crashes_df$M_HARM %in% c(98, 99)                         ~ NA_character_,
    crashes_df$M_HARM == 8                                   ~ "pedestrian",
    crashes_df$M_HARM %in% 1:7                               ~ "non_collision",
    crashes_df$M_HARM %in% c(12, 14, 45, 51, 54, 55)         ~ "motor_vehicle",
    crashes_df$M_HARM %in% c(9, 10, 11, 15, 49, 74)          ~ "other_non_motorist",
    crashes_df$M_HARM %in% c(16, 17, 18, 73, 91)             ~ "falling_object",
    crashes_df$M_HARM %in% c(19, 20, 21, 23:26, 30:35,
                       38:43, 46, 48, 50, 52, 53,
                       57:59, 72, 93, 44)              ~ "fixed_object",
    TRUE                                                ~ NA_character_
    )
  
  # 3 categories
  crashes_df$P_CRASH1 <- dplyr::case_when(
    crashes_df$P_CRASH1 %in% c(0, 99)                    ~ NA_character_,
    crashes_df$P_CRASH1 %in% c(1, 2, 3)                  ~ "straight_travel",
    crashes_df$P_CRASH1 %in% c(4, 5, 7, 8, 9)            ~ "stopped_or_parking",
    crashes_df$P_CRASH1 %in% c(6, 10, 11, 12, 13, 
                          14, 15, 16, 17, 98)       ~ "complex_maneuver",
    TRUE                                            ~ NA_character_
    )
  
  # 3 categories
  crashes_df$VE_FORMS <- dplyr::case_when(
    crashes_df$VE_FORMS == 1  ~ "single",
    crashes_df$VE_FORMS == 2  ~ "two",
    crashes_df$VE_FORMS >= 3  ~ "multiple",
    TRUE                ~ NA_character_
  )
  
  # 5 categories
  crashes_df$ROUTE <- dplyr::case_when(
    crashes_df$ROUTE == 99                   ~ NA_character_,
    crashes_df$ROUTE %in% c(1, 11)           ~ "interstate",
    crashes_df$ROUTE %in% c(2, 3)            ~ "highway",
    crashes_df$ROUTE %in% c(4, 5, 6, 12)     ~ "local",          # county/township/municipal
    crashes_df$ROUTE %in% c(0, 96)           ~ "unsigned",       # not signed + not in inventory
    crashes_df$ROUTE %in% c(10, 13, 95)      ~ "other",
    TRUE                               ~ NA_character_
  )
  
  # 6 categories
  crashes_df$WEATHER <- dplyr::case_when(
    crashes_df$WEATHER %in% c(98, 99)       ~ NA_character_,
    crashes_df$WEATHER == 1                  ~ "clear",
    crashes_df$WEATHER == 10                 ~ "cloudy",
    crashes_df$WEATHER == 2                  ~ "rain",
    crashes_df$WEATHER %in% c(3, 4, 11, 12) ~ "frozen_precip",  # snow/sleet/freezing rain
    crashes_df$WEATHER == 5                  ~ "fog",
    crashes_df$WEATHER %in% c(6, 7, 8)      ~ "other",
    TRUE                               ~ NA_character_
  )
  
  crashes_df$PREV_ACC <- dplyr::case_when(
    crashes_df$PREV_ACC %in% c(98, 99, 998) ~ NA_character_,
    crashes_df$PREV_ACC == 0                 ~ "none",
    crashes_df$PREV_ACC == 1                 ~ "one",
    crashes_df$PREV_ACC >= 2                 ~ "multiple",
    TRUE                               ~ NA_character_
  )
  
  # 4 categories
  crashes_df$VPAVETYP <- dplyr::case_when(
    crashes_df$VPAVETYP %in% c(8, 9)      ~ NA_character_,
    crashes_df$VPAVETYP %in% c(1, 2, 3)   ~ "paved",        # concrete/asphalt/brick
    crashes_df$VPAVETYP %in% c(4, 5)      ~ "unpaved",       # gravel/dirt
    crashes_df$VPAVETYP == 0               ~ "non_trafficway",
    crashes_df$VPAVETYP == 7               ~ "other",
    TRUE                             ~ NA_character_
  )
  
  # 5 categories
  crashes_df$VSURCOND <- dplyr::case_when(
    crashes_df$VSURCOND %in% c(98, 99)     ~ NA_character_,  # missing
    crashes_df$VSURCOND == 1                ~ "dry",
    crashes_df$VSURCOND == 2                ~ "wet",
    crashes_df$VSURCOND %in% c(3, 4, 10)   ~ "frozen",       # snow/ice/slush
    crashes_df$VSURCOND %in% c(0,5,6,7,8,11) ~ "other",        # sand/water/oil/mud etc.
    TRUE                               ~ NA_character_
  )
  
  # 5 categories
  crashes_df$LGT_COND <- dplyr::case_when(
    crashes_df$LGT_COND %in% c(8, 9)  ~ NA_character_,
    crashes_df$LGT_COND == 1           ~ "daylight",
    crashes_df$LGT_COND == 2           ~ "dark_unlit",
    crashes_df$LGT_COND == 3           ~ "dark_lighted",
    crashes_df$LGT_COND %in% c(4, 5)  ~ "dawn_dusk",
    crashes_df$LGT_COND %in% c(6, 7)  ~ "dark_unknown",  # merge other into dark_unknown
    TRUE                          ~ NA_character_
  )
  
  return(crashes_df)
}
```

### Missingness to NA
```{r}
missNA <- function(crashes_df)
{
  # AGE; 4 categories
  crashes_df$AGE[crashes_df$AGE %in% c(998, 999)] <- NA
  
  # RUR_URB; 3 categories
  crashes_df$RUR_URB[crashes_df$RUR_URB %in% c(8, 9)] <- NA
  
  # RELJCT1; binary
  crashes_df$RELJCT1[crashes_df$RELJCT1 %in% c(8, 9)] <- NA
  
  # MECISSUE; binary
  crashes_df$MECISSUE[crashes_df$MECISSUE == 2] <- NA
  
  # VIS; binary
  crashes_df$VIS[crashes_df$VIS %in% c(2, 3)] <- NA
  
  # DR_AGE; 4 categories
  crashes_df$DR_AGE[crashes_df$DR_AGE %in% c(998, 999)] <- NA
  
  # DR_SEX; binary
  crashes_df$DR_SEX[crashes_df$DR_SEX %in% c(8, 9)] <- NA
  
  return(crashes_df)
}
```

### Bin Ordinal
```{r}
bin <- function(crashes_df)
{
  crashes_df$AGE <- cut(crashes_df$AGE,
                breaks = c(-1, 15, 25, 64, Inf),
                labels = c("child", "young_adult", "adult", "elderly"),
                right  = TRUE)

  crashes_df$DR_AGE <- cut(crashes_df$DR_AGE,
                  breaks = c(-1, 15, 25, 64, Inf),
                  labels = c("child", "young_adult", "adult", "elderly"),
                  right  = TRUE)
  
  return(crashes_df)
}
```

### Imputation
```{r}
impute <- function(crashes_df)
{
  # Run multiple imputation
  # m = number of imputed datasets (5 is standard)
  # method = "cart" works well for mixed variable types
  imputed <- mice(crashes_df,
                  m         = 1,
                  method    = "cart",
                  seed      = 42,
                  maxit     = 5,      # proper cycling, just one dataset
                  printFlag = TRUE)
  
  # Extract one complete dataset for RFCI
  # (for causal discovery we use a single completed dataset)
  complete(imputed, action = 1)
}
```

## Drop unused variables
```{r}
finalDrops <- function(crashes_df)
{
  # Drops variables used for construction, outcome leakage, and non-predictive columns
  # ====================================================================================
  # Inputs
  # - crashes_df: full joined and derived data frame
  # Outputs
  # - data frame with only analysis-ready predictor columns retained
  
  crashes_df %>%
    # Dropping Predictors of outcome
    select(-PER_TYP, -INJ_SEV)%>%
    # Dropping Variables used to construct other variables but we don't want
    select(-MONTH, -DAY, -HOUR, -MINUTE, -DEATH_YR, -DEATH_MO,
           -DEATH_DA, -DEATH_HR, -DEATH_MN, -DEATH_TM, -ARR_HOUR,
           -ARR_MIN, -LAG_HRS, -LAG_MINS) %>%
    # Dropping Non-predictive variables for the non-driver
    select(-SEX)
}
```

## Factorize
```{r}
factorize <- function(crashes_df) {
  crashes_df %>%
    mutate(across(where(is.character), as.factor))
}
```

## Total Function Call
Function to call
```{r}
processYears <- function(fileLoc, startYear, endYear) {
  # Full pipeline from raw CSVs to a clean analysis ready data frame
  # =================================================================
  # Inputs
  # - fileLoc:   path to the project data folder
  # - startYear: first year to import as integer
  # - endYear:   last year to import as integer
  # Outputs
  # - single clean data frame spanning all years ready for analysis

  files <- c("accident.csv", "PERSON.csv", "VEHICLE.csv", "distract.csv",
             "factor.csv", "nmcrash.csv", "cevent.csv", "vision.csv")
  
  merging_Vars <- c("ST_CASE", "VEH_NO", "PER_NO")
  accident_keep <- c("VE_FORMS", "MONTH", "DAY", "YEAR", "HOUR", "MINUTE",
                     "DAY_WEEK", "ROUTE", "RUR_URB", "RELJCT1", "LGT_COND",
                     "WEATHER", "ARR_HOUR", "ARR_MIN", "ROAD_FNC")
  vehicle_keep <- c("TRAV_SP", "DR_DRINK", "PREV_ACC", "VSPD_LIM", "VPAVETYP",
                    "VSURCOND", "P_CRASH1", "M_HARM")
  person_keep <- c("AGE", "SEX", "PER_TYP", "INJ_SEV", "DRUGS", "DEATH_MO",
                   "DEATH_DA", "DEATH_YR", "DEATH_TM", "DEATH_HR", "DEATH_MN",
                   "LAG_HRS", "LAG_MINS")
  cevenet_keep <- c("EVENTNUM", "VNUMBER1")
  vision_keep <- c("VISION", "MVISOBSC")
  fact_keep <- c("VEHICLECC", "MFACTOR")
  dist_keep <- c("DRDISTRACT", "MDRDSTRD")
  NMCC_keep <- c("NMCC", "MTM_CRSH")
  
  # Map file names to their keep variables
  var_map <- list(
  "accident.csv" = accident_keep,
  "VEHICLE.csv"  = vehicle_keep,
  "PERSON.csv"   = person_keep,
  "factor.csv"   = fact_keep,
  "distract.csv" = dist_keep,
  "nmcrash.csv"  = NMCC_keep,
  "vision.csv"   = vision_keep,
  "cevent.csv"   = cevenet_keep
  )

  # Import and trim all years
  trimmed_flat <- importYears(startYear, endYear, files, fileLoc) 
  trimmed_flat <- trimAllYears(trimmed_flat, startYear, endYear, var_map, merging_Vars)

  # Process, join within year, and stack all years
  all_years <- bind_rows(lapply(trimmed_flat, mergeYear))

  # Post-join derived variables and final column drops
  all_years <- all_years %>%
    addDerivedVariables()%>%
    missNA() %>%
    bin() %>%
    coarsen() %>%
    finalDrops() %>%
    factorize() %>%
    impute()
}
```

# Data Possessing
```{r}
fileLoc <- "../566 Project Data"

crashes2012to24 <- processYears(fileLoc, 2012, 2024)
```

# Export
```{r}
write.csv(crashes2012to24, "../566 Project Data/crashes2012to24.csv")
```